---
title: "Using securetools with agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using securetools with agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

An agent decides for itself when to call a tool and with what arguments.
Its choices depend on everything it has read, including text an attacker
may have planted. So any tool you give it can end up being called in ways
you didn't plan for, and the tool itself has to enforce the limits.

This vignette shows securetools inside
[orchestr](https://github.com/ian-flores/orchestr) agents. Each `tool_*()`
function returns a `securer::securer_tool()`, and orchestr's `agent()`
turns it into an ellmer tool when you set `secure = TRUE`. The limits
(allowed folders, allowed tables, call caps) come with the tool, so you
don't write any checking code yourself.

The examples cover a single agent in a ReAct loop, a supervisor with
specialist workers, call limits, and custom tools next to the built-in
ones. For the tools themselves, see `vignette("securetools")`. For orchestr,
see `vignette("quickstart", package = "orchestr")`.

## Setup

```{r setup}
library(securetools)
library(orchestr)
library(ellmer)

# Set your LLM provider API key
Sys.setenv(ANTHROPIC_API_KEY = "your-key-here")
```

## A ReAct agent with tools

ReAct (reason, then act) is the usual loop for an agent with tools. The
model reads the task, decides what to do, calls a tool, looks at the
result, and repeats until it has an answer. Every turn of the loop is
another tool call, and an agent that is confused or has been manipulated
can keep going for a long time.

With securetools, every call is checked in the parent R process before the
tool does anything. The model can ask for whatever it likes; the tool only
does what its limits allow.

Here is how one turn of the loop runs:

```{=html}
<figure class="hairline">
<svg role="img" aria-label="Reason, act, validate, execute, observe, then loop until done and return the answer" viewBox="0 0 900 270" xmlns="http://www.w3.org/2000/svg"><defs>
<marker id="rx-arrow" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
<path d="M1 1L9 5L1 9" fill="none" stroke="#2b1f12" stroke-width="1"/></marker>
<pattern id="rx-hatch" width="6" height="6" patternUnits="userSpaceOnUse" patternTransform="rotate(45)">
<line x1="0" y1="0" x2="0" y2="6" stroke="#bf5a36" stroke-width="0.6" opacity="0.55"/></pattern>
</defs><rect x="20" y="112" width="120" height="64" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="80.0" y="134.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >REASON</text><text x="80.0" y="148.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the model decides</text><text x="80.0" y="161.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >what to do next</text><path d="M140 144.0H176" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)"/><rect x="178" y="112" width="120" height="64" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="238.0" y="134.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >ACT</text><text x="238.0" y="148.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >it asks for</text><text x="238.0" y="161.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >a tool call</text><path d="M298 144.0H334" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)"/><rect x="336" y="102" width="150" height="84" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="411.0" y="127.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >VALIDATE</text><text x="411.0" y="141.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >rate limit</text><text x="411.0" y="154.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >allow-list</text><text x="411.0" y="167.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >path check</text><path d="M486 144.0H522" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)"/><rect x="524" y="112" width="120" height="64" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="584.0" y="141.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >EXECUTE</text><text x="584.0" y="155.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the tool runs</text><path d="M644 144.0H680" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)"/><rect x="682" y="112" width="120" height="64" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="742.0" y="134.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >OBSERVE</text><text x="742.0" y="148.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the result goes</text><text x="742.0" y="161.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >back to the model</text><path d="M742 176V250H80V178" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)" stroke-dasharray="3 3"/><text x="411" y="244" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="2" >LOOP</text><path d="M80 112V44H676" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#rx-arrow)"/><text x="90" y="36" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.4" >done</text><rect x="678" y="24" width="200" height="40" fill="none" stroke="#bf5a36" stroke-width="0.75"/><text x="778.0" y="48.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#bf5a36" text-anchor="middle" font-weight="700" letter-spacing="1.2" >RETURN THE ANSWER</text><path d="M20 201V206H298V201" fill="none" stroke="#cdb48a" stroke-width="0.75"/><text x="159.0" y="221" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.6" >THE MODEL</text><path d="M336 201V206H486V201" fill="none" stroke="#cdb48a" stroke-width="0.75"/><text x="411.0" y="221" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.6" >YOUR R SESSION</text><path d="M524 201V206H644V201" fill="none" stroke="#cdb48a" stroke-width="0.75"/><text x="584.0" y="221" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.6" >THE TOOL</text></svg>
<figcaption>Fig. 1 &middot; One turn of the ReAct loop, with securetools checks in the middle</figcaption>
</figure>
```

Pass the tools to `agent()` with `secure = TRUE` so the agent's code runs
in a securer session.

```{r react-basic}
# Create security-scoped tools
calc <- tool_calculator()
reader <- tool_read_file(allowed_dirs = "/path/to/project/data")

# Build an agent with tools and secure execution
analyst <- agent(
  "analyst",
  chat = chat_anthropic(
    system_prompt = "You are a data analyst. Use your tools to answer questions."
  ),
  tools = list(calc, reader),
  secure = TRUE
)

# Wrap in a ReAct graph for state management
graph <- react_graph(analyst)

result <- graph$invoke(list(messages = list(
  "Read the file sales.csv from the data directory and calculate the total revenue."
)))
```

With `secure = TRUE`, orchestr starts a `SecureSession` and converts each
`securer_tool` into an ellmer tool. The path checks, the calculator's
function list, and the call limits all run in the parent process.

## A supervisor with specialist workers

Another option is to split the tools across several agents, so each one
has only what it needs. A supervisor agent reads the request and hands it
to the right worker. The supervisor has no tools of its own.

Each worker gets its own `SecureSession`, with its own allowed folders,
allowed domains, and call limits. A file worker that gets hijacked still
can't make web requests, because it was never given that tool. The
supervisor only sees the workers' answers, not the raw tool output. You can
also set limits per role, for example a generous call limit for the data
worker and a tight one for the worker that writes files.

```{=html}
<figure class="hairline">
<svg role="img" aria-label="A supervisor routes a request to a file worker, a data worker, or a research worker" viewBox="0 0 900 262" xmlns="http://www.w3.org/2000/svg"><defs>
<marker id="sv-arrow" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
<path d="M1 1L9 5L1 9" fill="none" stroke="#2b1f12" stroke-width="1"/></marker>
<pattern id="sv-hatch" width="6" height="6" patternUnits="userSpaceOnUse" patternTransform="rotate(45)">
<line x1="0" y1="0" x2="0" y2="6" stroke="#bf5a36" stroke-width="0.6" opacity="0.55"/></pattern>
</defs><text x="450" y="22" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1" >&#8220;READ SALES.CSV, COMPUTE MEAN REVENUE&#8221;</text><path d="M450 30V48" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#sv-arrow)"/><rect x="340" y="50" width="220" height="56" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="450.0" y="68.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >SUPERVISOR</text><text x="450.0" y="82.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >routes the request</text><text x="450.0" y="95.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >has no tools of its own</text><path d="M450 106V128" fill="none" stroke="#2b1f12" stroke-width="0.75"/><path d="M170 128H730" fill="none" stroke="#2b1f12" stroke-width="0.75"/><path d="M170 128V158" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#sv-arrow)"/><rect x="52" y="152" width="236" height="96" fill="none" stroke="#cdb48a" stroke-width="0.75" stroke-dasharray="3 3"/><rect x="60" y="160" width="220" height="66" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="170.0" y="183.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >FILE WORKER</text><text x="170.0" y="197.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >read_file</text><text x="170.0" y="210.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >write_file</text><text x="170" y="240" font-family="Space Mono, ui-monospace, monospace" font-size="8" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.4" >OWN SECURESESSION</text><path d="M450 128V158" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#sv-arrow)"/><rect x="332" y="152" width="236" height="96" fill="none" stroke="#cdb48a" stroke-width="0.75" stroke-dasharray="3 3"/><rect x="340" y="160" width="220" height="66" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="450.0" y="183.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >DATA WORKER</text><text x="450.0" y="197.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >calculator</text><text x="450.0" y="210.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >data_profile</text><text x="450" y="240" font-family="Space Mono, ui-monospace, monospace" font-size="8" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.4" >OWN SECURESESSION</text><path d="M730 128V158" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#sv-arrow)"/><rect x="612" y="152" width="236" height="96" fill="none" stroke="#cdb48a" stroke-width="0.75" stroke-dasharray="3 3"/><rect x="620" y="160" width="220" height="66" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="730.0" y="190.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >RESEARCH WORKER</text><text x="730.0" y="204.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >fetch_url</text><text x="730" y="240" font-family="Space Mono, ui-monospace, monospace" font-size="8" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.4" >OWN SECURESESSION</text></svg>
<figcaption>Fig. 2 &middot; Each worker gets its own session, rate limits, and allow-lists</figcaption>
</figure>
```

Here is a supervisor with a data worker and a file worker:

```{r supervisor}
# Data agent: calculation and profiling
data_agent <- agent(
  "data_specialist",
  chat = chat_anthropic(
    system_prompt = paste(
      "You are a data specialist.",
      "Use the calculator for arithmetic and the profiler for data summaries."
    )
  ),
  tools = list(
    tool_calculator(),
    tool_data_profile(max_rows = 50000)
  ),
  secure = TRUE
)

# File agent: reading and writing
file_agent <- agent(
  "file_specialist",
  chat = chat_anthropic(
    system_prompt = paste(
      "You are a file specialist.",
      "Read and write files as requested.",
      "Always specify format = 'auto' when reading."
    )
  ),
  tools = list(
    tool_read_file(allowed_dirs = "/path/to/project/data"),
    tool_write_file(allowed_dirs = "/path/to/project/output")
  ),
  secure = TRUE
)

# Supervisor routes between specialists
supervisor <- agent(
  "supervisor",
  chat = chat_anthropic(
    system_prompt = paste(
      "You coordinate a team.",
      "Route data questions to the data specialist",
      "and file operations to the file specialist."
    )
  )
)

graph <- supervisor_graph(
  supervisor = supervisor,
  workers = list(
    data_specialist = data_agent,
    file_specialist = file_agent
  )
)

result <- graph$invoke(list(messages = list(
  "Read sales.csv, then calculate the mean of the revenue column."
)))
```

The supervisor routes with a `route` tool that `supervisor_graph()` adds
for it.

## Call limits in agent loops

In an agent loop, the model decides how many times to call a tool. An agent
that misreads a task might call the calculator 500 times to "check" its
answer. A research agent might keep following links and hammer an API. Each
of those calls costs tokens and time.

securetools has two kinds of limit:

- `max_calls` caps the total number of calls to a tool. Once it's reached,
  every later call fails. This is the one that stops a runaway loop.
- `max_calls_per_minute` (only on `tool_fetch_url()`) caps how often the
  tool can be called. You might allow 1000 calls in total but only 10 a
  minute, so an external service isn't flooded.

When a limit is reached, the tool raises an error. ellmer passes the error
back to the model as the tool's result, so the agent can stop or, for
example, summarize what it already has instead of fetching more.

```{r rate-limiting}
# Cap the calculator at 50 calls per agent session
calc <- tool_calculator(max_calls = 50)

# URL fetch with both lifetime and per-minute limits
fetcher <- tool_fetch_url(
  allowed_domains = c("api.github.com"),
  max_calls = 100,
  max_calls_per_minute = 10
)

researcher <- agent(
  "researcher",
  chat = chat_anthropic(
    system_prompt = "You fetch data from APIs and analyze results."
  ),
  tools = list(calc, fetcher),
  secure = TRUE
)

graph <- react_graph(researcher)
```

`tool_fetch_url()` can't fetch HTTPS URLs at the moment (see
`vignette("securetools")`), so this agent would get an error back from
`https://api.github.com` until that is fixed.

## Custom tools next to securetools

Most agents need something securetools doesn't have. You can put your own
`securer::securer_tool()` definitions in the same agent as the built-in
ones.

```{r custom-tool}
# A custom tool alongside securetools
timestamp_tool <- securer::securer_tool(
  name = "timestamp",
  description = "Return the current UTC timestamp.",
  fn = function() {
    format(Sys.time(), tz = "UTC", usetz = TRUE)
  },
  args = list()
)

# Mix custom + securetools
assistant <- agent(
  "assistant",
  chat = chat_anthropic(
    system_prompt = "You help with data tasks and can check the current time."
  ),
  tools = list(
    tool_calculator(),
    tool_read_file(allowed_dirs = "/path/to/data"),
    timestamp_tool
  ),
  secure = TRUE
)

graph <- react_graph(assistant)

result <- graph$invoke(list(messages = list(
  "What time is it? Also, what is 2^10?"
)))
```

A custom tool's `fn` runs in the parent process, like the securetools
ones. The difference is that it only has the checks you write into it.

## Next steps

- `vignette("securetools")`: the tools and what each one checks
- `vignette("quickstart", package = "orchestr")`: agent and graph basics
- `vignette("ellmer-integration", package = "securer")`: connecting securer to ellmer directly
